Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-11-09 notebooks/expected_and_actual_counts_ratio.Rmd
Add ability to compare clades using amino acids

library(data.table)
library(ggplot2)
library(ggforce)
library(stringr)
library(gtools)
```

@@

# To avoid infinities when calculating fold-change, add a pseudocount to the observed and expected counts.
pseudo <- 0.5
```

@@

@@

@@

Describe the filtered data set.

```{r describe_data}
# How many rows? How many columns?
dim(filtered_SARS_mutation)

# How many nucleotide sites?
length(unique(filtered_SARS_mutation$nt_site))
```

@@

[INSERT] information about the plot. Plot to visualize data.

```{r stacked_bar, echo=FALSE}
# Add column with the mutation type (clade founder.mutation).
add_mutation <- filtered_SARS_mutation
add_mutation$mutation_type <- paste(str_sub(add_mutation$nt_mutation,1,1), ".", str_sub(add_mutation$nt_mutation,-1,-1), sep="")

# Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
ggplot(add_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
  geom_bar(position="stack", stat="identity")+
  theme_bw()+
  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

[INSERT] information about the plot. Plot to visualize data.

```{r nucleotide_site, echo=FALSE}
ggplot(add_mutation, aes(color=mutation_type, y=log(actual_count/expected_count), x=nt_site))+
  geom_point()+
  theme_bw()+
  labs(x="", y="")
```

@@

```{r subset_by_clade}
# Create a subset of the data using only the clades that are going to be compared.
filtered_clades <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]

# Remove columns that were already used to filter the data or than contain redundant information.
filtered_clades$subset <- NULL
filtered_clades$exclude <- NULL
filtered_clades$nt_site <- NULL
```

@@

# Append 'clade1' or 'clade2' to the column names so that the two data frames can be combined with easy to interpret columns.
colnames(clade1) <- paste("c1", colnames(clade1), sep="_")
colnames(clade2) <- paste("c2", colnames(clade2), sep="_")

# Modify nt_mutation, clade_founder_nt and clade to make merging easier and make the data frame more readable.
{colnames(clade1)[which(colnames(clade1)=="c1_nt_mutation")] <- "nt_mutation"
colnames(clade2)[which(colnames(clade2)=="c2_nt_mutation")] <- "nt_mutation"
colnames(clade1)[which(colnames(clade1)=="c1_clade_founder_nt")] <- "clade_founder_nt"
colnames(clade2)[which(colnames(clade2)=="c2_clade_founder_nt")] <- "clade_founder_nt"
colnames(clade1)[which(colnames(clade1)=="c1_clade")] <- "clade1"
colnames(clade2)[which(colnames(clade2)=="c2_clade")] <- "clade2"}

# Merge the two data frames by mutation so that each row represents one comparison.
comparisons <- merge(clade1, clade2, by=c("nt_mutation","clade_founder_nt"))
```

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["c1_actual_count"]),as.numeric(c["c1_rounded_expected"])),c(as.numeric(c["c2_actual_count"]),as.numeric(c["c2_rounded_expected"]))))$p.value}))

# Save and load variable because calculating p-values may take a long time.
#save(p_values, file=paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))
load(paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))
```

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots.
comparison_p_values$fold_change <- log2(comparison_p_values$c1_observed_expected/comparison_p_values$c2_observed_expected)
```

@@

```{r volcano_plot, echo=FALSE}
# Plot the uncorrected p-values.
ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
  geom_point(alpha=0.5)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  theme_bw()

# After dropping p = 1, plot FDR corrected p-values (color by nucleotide clade founder and label by mutation).
volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=nt_mutation, color=clade_founder_nt))+
  geom_point()+
  geom_text(color="black", size=2, nudge_y=0.2)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  scale_color_brewer(palette="Set2")+
  theme_bw()
```

@@

```{r verify_analyses}
# Plot the corrected p-values for synonymous and non-synonymous mutations and compare to a significance value of 0.05.
verify_analyses <- comparison_p_values
verify_analyses$synonymous <- ifelse(verify_analyses$c1_synonymous=="TRUE" & verify_analyses$c2_synonymous=="TRUE", "TRUE", "FALSE")
verify_analyses$p_one <- ifelse(verify_analyses$p_value_corrected==1, "P=1", "P<1")

>>>>

library(data.table)
library(ggplot2)
library(ggforce)
library(stringr)
library(gtools)

`%notin%` <- Negate(`%in%`)
```

@@

# To avoid infinities when calculating fold-change, add a pseudocount to the observed and expected counts.
pseudo <- 0.5

option <- "aa" # nt or aa
merge_columns <- c("nt_mutation","nt_site","clade_founder_nt","gene","codon_position")
merge_columns <- c("aa_mutation","gene")
```

@@

With the data set, both nucleotide and amino acid mutations can be compared between clades. However, there is no column identifying the amino acid mutation (i.e., starting amino acid, codon position along the genome and mutated amino acid). Create a column with this information using the nucleotide site, codon position (i.e., 1, 2, or 3), clade founder amino acid and mutant amino acid columns.

```{r aa_mutation}
# Create a list with the starting nucleotide site for an amino acid for all rows in the data set.
starting_nt <- SARS_mutation$nt_site-as.numeric(str_sub(SARS_mutation$codon_position,1,1))+1

# Create column with amino acid mutation (e.g., M.266.268.L)
SARS_mutation$aa_mutation <- paste(str_sub(SARS_mutation$clade_founder_aa,1,1), starting_nt, starting_nt+2, str_sub(SARS_mutation$mutant_aa,1,1), sep=".")
```

@@

```{r clean_columns}
# Remove columns that were already used to filter the data.
filtered_SARS_mutation$subset <- NULL
filtered_SARS_mutation$exclude <- NULL
```

@@

Look at a summary of the data.\
-What are the data types for each column?\
-Is there a similar number of TRUEs and FALSEs for logical variables?\
-Describe the numeric variables (e.g., minimum value).\

```{r summary_data}
summary(filtered_SARS_mutation)
```

Take a closer look at how many nucleotide sites and the proportion of mutations in the data set.

```{r describe_data}
# How many nucleotide sites?
length(unique(filtered_SARS_mutation$nt_site))

# What percentage of the data set is each type of mutation? (e.g., A -> T)
mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
(table(mutation_type)/length(mutation_type))*100
```

@@

The four nucleotides (e.g., A) and their possible mutations (e.g., A -> T) may be under different directions and degrees of selection between clades leading to an unequal distribution of nucleotides in the genome. Look at how many mutations (normalized by the expected count) there are for each clade.

```{r stacked_nucleotide, echo=FALSE}
stacked_nt <- filtered_SARS_mutation
stacked_nt$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")

ggplot(stacked_nt, aes(x=clade, y=actual_count/expected_count, fill=mutation_type))+
  geom_bar(position="stack", stat="identity")+
  scale_fill_brewer(palette="Paired")+
  theme_bw()+
  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

Amino acids occur at different proportions throughout the SARS-CoV-2 genome and between phylogenetic clades. [INSERT]

```{r stacked_amino_acid, echo=FALSE}
stacked_aa <- filtered_SARS_mutation
stacked_aa$clade_founder <- str_sub(filtered_SARS_mutation$clade_founder_aa,1,1)
stacked_aa$mutant <- str_sub(filtered_SARS_mutation$mutant_aa,1,1)

ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant))+
  geom_bar(position="stack", stat="identity")+
  facet_wrap(~clade_founder, ncol=7)+
  theme_bw()+
  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
```

@@

```{r subset_by_clade}
# Create a subset of the data using only the clades that are going to be compared.
filtered_clades <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
```

@@

# Append 'clade1' or 'clade2' to the column names so that the two data frames can be combined with easy to interpret columns.
colnames(clade1)[colnames(clade1) %notin% merge_columns] <- paste(colnames(clade1)[colnames(clade1) %notin% merge_columns], "1", sep="_")
colnames(clade2)[colnames(clade2) %notin% merge_columns] <- paste(colnames(clade2)[colnames(clade2) %notin% merge_columns], "2", sep="_")

# Merge the two data frames by mutation so that each row represents one comparison.
comparisons <- merge(clade1, clade2, by=merge_columns)
```

@@

```{r contingency_table}
# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
#p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["actual_count_1"]),as.numeric(c["rounded_expected_1"])),c(as.numeric(c["actual_count_2"]),as.numeric(c["rounded_expected_2"]))))$p.value}))

# Save and load variable because calculating p-values may take a long time.
#save(p_values, file=paste("pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
load(paste("pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
```

@@

```{r fold-change}
# Calculate fold change for plotting volcano plots.
comparison_p_values$fold_change <- log2(comparison_p_values$observed_expected_1/comparison_p_values$observed_expected_2)
```

@@

```{r volcano_all, echo=FALSE}
# Plot the uncorrected p-values.
ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
  geom_point(alpha=0.5)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  theme_bw()
```

```{r volcano_corrected, echo=FALSE}
# After dropping p = 1, plot FDR corrected p-values (color by nucleotide clade founder and label by mutation).
volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(paste(option, "mutation", sep="_")), color=str_sub(get(paste(option, "mutation", sep="_")),1,1)))+
  geom_point()+
  geom_text(color="black", size=2, nudge_y=0.2)+
  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
  #scale_color_brewer(palette="Set2", name="mutation")+
  theme_bw()
```

@@

```{r verify_analyses}
# Plot the corrected p-values for synonymous and non-synonymous mutations and compare to a significance value of 0.05.
verify_analyses <- comparison_p_values
verify_analyses$synonymous <- ifelse(verify_analyses$synonymous_1=="TRUE" & verify_analyses$synonymous_2=="TRUE", "TRUE", "FALSE")
verify_analyses$p_one <- ifelse(verify_analyses$p_value_corrected==1, "P=1", "P<1")
